---
title: "p8105_hw2_xw2972"
author: "Xiwen Wu"
output: github_document
date: 2024-10-15
---

Question 1
This dataset contains 2,595,176 observations and 7 variables, with weather data recorded across different stations in New York State. The variables are: id, date, prcp(precipitation), snow(snow fall), snwd(snow depth), tmax, tmin. A significant portion of the dataset contains missing values, particularly in the tmax, tmin, prcp , snow, and snwd. This might occur due to incomplete records, or days without precipitation and snow. These missing values could limit the scope of analysis.
For snowfall, what are the most commonly observed values is 0, this is due to typical weather patterns in New York State, where days without snow are frequent.
The temperatures are significantly higher in July compared to January. There is one station that has average max temperature lower than 10 degrees in January, which seems like an outlier. 

```{r}
library(tidyverse)
library(ggridges)
library(p8105.datasets)
#a
data("ny_noaa")
ny_noaa = 
    janitor::clean_names(ny_noaa)
ny_noaa <- ny_noaa %>%
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    prcp = prcp / 10,               
    tmax = as.numeric(tmax) / 10,  
    tmin = as.numeric(tmin) / 10,   
    snow = snow
  )%>%
  select(id, year, month, day, prcp, snow, snwd, tmax, tmin)

snowfall_summary <- ny_noaa %>%
  group_by(snow) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
head(snowfall_summary)

#b
avg_tmax <- ny_noaa %>%
  filter(month == 1 | month == 7) %>%
  group_by(id, month, year) %>%
  summarize(avg_tmax = mean(tmax, na.rm = TRUE), .groups = 'drop') %>%
  mutate(month = ifelse(month == 1, "January", "July")) %>%
  group_by(id, month) %>%
  summarize(avg_tmax = mean(avg_tmax, na.rm = TRUE), .groups = 'drop')

avg_tmax <- avg_tmax %>%
  filter(!is.na(avg_tmax))

ggplot(avg_tmax, aes(x = id, y = avg_tmax)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_grid(~ month) +
  labs(title = "Average Max Temperature in January and July by Station",
       x = "Stations",
       y = "Average Max Temperature (°C)") +
  theme(axis.text.x = element_blank())

#c1
par(mfrow = c(1, 2)) 

tmax_density <- density(ny_noaa$tmax, na.rm = TRUE)
tmin_density <- density(ny_noaa$tmin, na.rm = TRUE)
plot(tmax_density, 
     main = "Density of Max Temperature",
     xlab = "Temperature (°C)",
     ylab = "Density",
     col = "orange", lwd = 2,
     xlim = range(c(tmax_density$x, tmin_density$x)), 
     ylim = c(0, max(c(tmax_density$y, tmin_density$y))))

plot(tmin_density, 
     main = "Density of Min Temperature",
     xlab = "Temperature (°C)",
     ylab = "Density",
     col = "blue", lwd = 2,
     xlim = range(c(tmax_density$x, tmin_density$x)), 
     ylim = c(0, max(c(tmax_density$y, tmin_density$y))))
par(mfrow = c(1, 1)) 

#c2
filtered_snowfall <- ny_noaa %>%
  filter(snow > 0 & snow < 100)

boxplot(snow ~ year, data = filtered_snowfall,
        main = "Distribution of Snowfall (0 < snow < 100) by Year",
        xlab = "Year",
        ylab = "Snowfall (inches)",
        col = "lightblue",
        outline = TRUE) 
```

Question 2
Chart A: Age Distribution by Education Level and Sex chart shows that younger individuals predominantly fall into the "More than high school" category, it suggests that educational attainment has improved over time.
Chart B: People who's education less than high school tends to have more total activity than people who's education more than high school during thier young age. 
Chart C: People with More than high school education show higher activity levels during usual commute time. 

```{r}
library(tidyverse)
library(ggridges)

demo_df <- read.csv("nhanes_covar.csv", skip = 4, na = c("NA", ".", ""))|>
  janitor::clean_names()

accel_df <- read.csv("nhanes_accel.csv", na = c("NA", ".", ""))|>
  janitor::clean_names()

data <- left_join(demo_df, accel_df, by = "seqn")

data <- data %>%
  mutate(
    MIMS = rowSums(select(., starts_with("min")), na.rm = TRUE),
    sex = recode(sex, `1` = "Male", `2` = "Female"),
    education = recode(education, `1` = "Less than high school", 
                                 `2` = "High school equivalent", 
                                 `3` = "More than high school")
  )%>% 
  filter(age >= 21) %>%
  drop_na()

table_education_sex <- data %>%
  group_by(education, sex) %>%
  summarise(n = n())

#Chart A
ggplot(data, aes(x = age, fill = as.factor(sex))) +
  geom_density(alpha = 0.5) +
  facet_grid(~education) +
  labs(title = "Age Distribution by Education Level and Sex", fill = "Sex")

#Chart B
data_total <- data %>%
  group_by(seqn) %>%
  summarise(
    total_activity = sum(MIMS, na.rm = TRUE), 
    age = mean(age), 
    sex = first(sex), 
    education = first(education)
  )

ggplot(data_total, aes(x = age, y = total_activity, color = as.factor(sex))) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(~education) +
  labs(title = "Total Activity vs Age by Education Level and Sex", color = "Sex")

#Chart C
data_long <- data %>%
  pivot_longer(cols = starts_with("min"),
               names_to = "time", 
               values_to = "activity")

data_long <- data_long %>%
  mutate(
    time = as.numeric(str_remove(time, "min")),  
    sex = as.factor(sex),                    
    education = as.factor(education)            
  )

ggplot(data_long, aes(x = time, y = activity, color = sex)) +
  geom_point(alpha = 0.3, size = 0.1) +  
  geom_smooth(se = FALSE) +  
  facet_grid(~education) +  
  labs(
    title = "24-Hour Activity Time Courses by Education Level and Sex", 
    color = "Sex", 
    x = "Time (Minutes)", 
    y = "Activity (MIMS)"
  ) +
  theme_minimal() 
```

Question 3
The combined dataset shows 99485 obs with 9 variables. 
The total_rides_table shows that members ride more than casual customers. People also tend to ride more in July compared to January in both 2020 and 2024.
According to Median Ride Duration by Day of the Week, Month, and Year chart, the median ride duration is longer in weekends than weekdays in both 2020 and 2024. 
According to Ride Duration Distribution by Membership, Month, and Bike Type (2024) chart, casual customers have longer ride duration.

```{r}
library(tidyverse)
library(ggridges)
jan_2020 <- read.csv("Jan 2020 Citi.csv",na = c("NA", ".", ""))|>
  janitor::clean_names()
jan_2024 <- read.csv("Jan 2024 Citi.csv",na = c("NA", ".", ""))|>
  janitor::clean_names()
july_2020 <- read.csv("July 2020 Citi.csv",na = c("NA", ".", ""))|>
  janitor::clean_names()
july_2024 <- read.csv("July 2024 Citi.csv",na = c("NA", ".", ""))|>
  janitor::clean_names()

jan_2020 <- jan_2020 %>%
  mutate(year = 2020, month = "January")
jan_2024 <- jan_2024 %>%
  mutate(year = 2024, month = "January")
july_2020 <- july_2020 %>%
  mutate(year = 2020, month = "July")
july_2024 <- july_2024 %>%
  mutate(year = 2024, month = "July")

citi_data <- bind_rows(jan_2020, jan_2024, july_2020, july_2024)

citi_data <- citi_data %>%
  mutate(
    rideable_type = as.factor(rideable_type),
    weekdays = as.factor(weekdays),
    member_casual = as.factor(member_casual),
    start_station_name = as.factor(start_station_name),
    end_station_name = as.factor(end_station_name),
    duration = as.numeric(duration)
  )

total_rides_table <- citi_data %>%
  group_by(year, month, member_casual) %>%
  summarise(total_rides = n()) %>%
  arrange(year, month)
print(total_rides_table)

top_stations_july_2024 <- citi_data %>%
  filter(year == 2024, month == "July") %>%
  group_by(start_station_name) %>%
  summarise(total_rides = n()) %>%
  arrange(desc(total_rides)) %>%
  slice(1:5)
print(top_stations_july_2024)

citi_data <- citi_data %>%
  mutate(weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))
median_duration_plot <- citi_data %>%
  group_by(year, month, weekdays) %>%
  summarise(median_duration = median(duration, na.rm = TRUE)) %>%
  ggplot(aes(x = weekdays, y = median_duration, color = month)) +
  geom_point() +
  facet_grid(~year) +
  labs(title = "Median Ride Duration by Day of the Week, Month, and Year", 
       x = "Day of the Week", y = "Median Duration (minutes)", color = "Month") +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 5)) 
print(median_duration_plot)

ride_duration_plot_2024 <- citi_data %>%
  filter(year == 2024) %>%
  ggplot(aes(x = duration, fill = rideable_type)) +
  geom_density(alpha = 0.5) +
  facet_grid(member_casual ~ month) +
  labs(title = "Ride Duration Distribution by Membership, Month, and Bike Type (2024)", 
       x = "Ride Duration (minutes)", y = "Density", fill = "Bike Type") +
  theme_minimal()
print(ride_duration_plot_2024)


```